Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-31 manu-draft.Rmd
no separate PCA and corr analysis for water quality

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.  Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

@@

Multivariate assessments of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)). The first principal component explained 36\%, 49\%, and 42\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.  These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.  The second and third principal components explained less of the variation among water quality parameters, such that 36\%, 20\%, and 34\% was explained by both axes for areas 1, 2, and 3, respectively.  The patterns explained by these axes were less clear, although physical drivers were partially explained in some cases (e.g., component 2 for area 2 explained a positive association for salinity and temperature, Figure \@ref(fig:pcacors)b).  Spearman rank correlations between the pairs of variables used for each of the PCAs showed significant associations.  Total nitrogen was positively correlated with chlorophyll and negatively associated with secchi depth for areas 1 and 2 ($p < 0.005$ and $p < 0.05$, respectively).  Dissolved oxygen was negatively associated with chl-a only in area 3 ($p < 0.005$), although the temporal progression likely changed over time in the other areas (e.g., increase followed by decrease).  Temperature generally had a positive association with chlorophyll ($p < 0.05$ for area 2, $p < 0.005$ for area 3).  pH was positively associated with chl-a in areas 1 and 2 ($p < 0.05$ for both).

@@

```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).  All variables were standardized to zero mean and unit variance prior to PCA. Variables with log-normal distribution were log-transformed prior to analysis.  Pearson correlations indicate the linear strength of assocation between pairs of variables. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
knitr::include_graphics(here('figs/pcacors.jpeg'))
```

>>>>

Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.  Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Further analyses were conducted to compare all data types together using PCA and correlation analyses, to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts. Data describing *K. brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.  Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.

@@

@@
